Choosing the Choice: Reflections on Modelling Decisions and Behaviour in Demographic Agent-Based Models

### 2017-03-29 article.tex
incorporated JB changes

In this research note, we discuss the challenges of choosing an appropriate model of choice in the context of decision making in agent-based demography. The argument is illustrated by examples related to demographic agent-based models, with focus on migration. In particular, in Section \ref{sec:context_matters} we elaborate on the role of the context of the decision processes. In Section \ref{auxiliary-factors-in-modelling-decisions}, we discuss some specific aspects of the mechanics of choice: time, heterogeneity of agents, as well as the choice between the choice models as such. Finally, in Section \ref{multi-model-approaches-and-modularity}, we conclude with a list of tentative recommendations for further work in the area, focusing on the mechanisms of decision making, multi-disciplinary approaches, a multi-model framework, and modularity of agent-based modelling.

@@

\citeauthor{McKenzie2013} also suggest that overweighting of negative experiences plays a role, which demonstrates the importance of considering how information is incorporated into the decision process. In this case the suggestion is that another cognitive bias, loss aversion, is a salient feature. This could be reflected by using a decision model explicitly incorporates loss aversion, for example Prospect Theory \citep{Kahneman1979}, or Cumulative Prospect Theory \citep{Tversky1992}. Both variants of Prospect Theory are intended to mimic human decision making, by distorting the perception of high and low probabilities towards certainty, and shifting the value of gains and losses such that large gains are progressively less significant, and losses more so. 
The wider implication is that social information flow may be distorted at the source, or through interpretation, and that agent-based models are uniquely suited to capture this.  We have focused here on descriptive theories of decision making, precisely because the aim is to model human behaviours as as they are manifested, rather than how they should be. Other approaches to decision models are available, for example heuristic methods \citep{Gigerenzer1996}, which emphasise the bounded rationality of human decision making, and Bayesian approaches, which address learning as a component of the decision process \citep{Robbins1964}. In an agent-based modelling context, \citet{Gray2016} have contrasted Bayesian, heuristic, and Cumulative Prospect Theory approaches in a simulation of alcohol misuse disclosure among pregnant women.

@@

Time is explicitly represented in almost all agent-based models, either in discrete or (pseudo) continuous form, in order to allow the dynamic unfolding of the modelled processes by `stepping' from one time-step to the next (or between time-ordered events in continuous-time discrete-event simulation). However, time often  does not enter into the decision making process of the simulated individuals. This results in the exclusion of a number of factors that may potentially be significant in eventual model outcomes. Firstly, individuals do not always make decisions instantaneously; as \cite{Willekens2017} describe elsewhere in this special issue, the theory of planned behaviour can be used to describe the gradual process of intention forming, influence-absorption, information gathering and planning that terminate in a migration event. This drawn-out decision making process allows the possibility that agents may change their minds. Secondly, past experiences affect decisions in the present. This brings up questions about how the passage of time affects memory and attitudes. Finally, agents differ in their preferences between rewards realised immediately and equivalent rewards realised at some point in the future.

@@

Unfortunately, this is not the limit of potential complications around time. A large number of other contextual factors make a difference to the extent of discounting. For instance, it has been found that smaller rewards are more heavily discounted than larger rewards; that people treat a reductions in time to rewards differently to equivalent delays; and future losses treated as more significant than equivalent gains. \citep{Read2000}.

@@

@@

Second, the existing methods of the statistical design of experiments (e.g. \citeauthor{Chaloner} \citeyear{Chaloner}) would allow for adopting a systematic approach to model design and data collection. In this way, following the suggestions of \citet{Courgeau}, the simulations would be built iteratively, by applying the experimental design principles to identifying the elements of the model which require additional empirical information, and then by collecting this information.  In this way, the important aspects of the decision process discussed above – context, time and heterogeneity – can be explicitly built into the model, through its specification or parameters. Then, they can be learnt about through repeated collection of information – including through experiments. The modularity of the model would help this process, as it would allow for fine-tuning the different aspects of the model independently, before they can be assembled together. An outline of the process is illustrated in Figure \ref{fig:approach}.

@@

As argued throughout this Special Issue, all these features, and more, make agent-based modelling a very attractive analytical proposition for demographers. The potential gains in understanding of the underlying population processes cannot be overstated, and further links connecting agent-based and statistical demography are yet to be explored (see for example \citet{Bijak2016}). That at present there is no methodological consensus on modelling agency and human decisions is not an obstacle, but rather a challenge to guide the further work in this area. Given that over the last 15 years agent-based models have slowly begun to enter the demographic mainstream \citep{Billari2003,VanBavel2016}, the gaps in knowledge in the existing approaches have become clearly visible. In our view one of the important gaps is related to the agency and decision making processes of the simulated agents, and the related aspects, some of which – context, time, and a variety of forms of decision making – are discussed above. Now is the time to initiate the discussion about choosing the choice.

>>>>

One of the most pervasive features of making decisions in the social realm is their uncertainty: the actors involved often do not -- and cannot -- know the full consequences of their decisions, nor even the full circumstances under which the decisions are made. However, in agent-based models, designed to reflect the complex features of social reality, uncertainty is typically being underplayed. As a result, the simulated agents lack psychological realism with respect to how they deal with ambiguity. Similarly, an analysis of additional layers of uncertainty, such as those related to model choice, model parameters, stochastic nature of the computer code, and so on, are often missing in model-building endavours. On the other hand, there is an acute need for bridging the gap between agent-based modelling and statistical inference, which can provide an appropriate language for dealing with uncertainty (see e.g. \citeauthor{Heard2015} \citeyear{Heard2015}).

In this paper, we discuss selected challenges of choosing an appropriate model of choice in the context of decision making in agent-based demography. The argument is illustrated by examples related to demographic agent-based models, with focus on migration. In particular, in Section \ref{sec:context_matters} we elaborate on the role of the context of the decision processes. In Section \ref{auxiliary-factors-in-modelling-decisions}, we discuss some specific aspects of the mechanics of choice: time, uncertainty, as well as the heterogeneity of agents. Finally, in Section \ref{multi-model-approaches-and-modularity}, we conclude with a list of tentative recommendations for further work in the area, focusing on the mechanisms of decision making, multi-disciplinary approaches, choice between the choice models as such, a multi-model framework, and modularity of agent-based modelling.

@@

\citeauthor{McKenzie2013} also suggest that overweighting of negative experiences plays a role, which demonstrates the importance of considering how information is incorporated into the decision process. In this case the suggestion is that another cognitive bias, loss aversion, is a salient feature. This could be reflected by using a decision model explicitly incorporates loss aversion, for example the Prospect Theory \citep{Kahneman1979}, or Cumulative Prospect Theory \citep{Tversky1992}. Both variants of Prospect Theory are intended to mimic human decision making, by distorting the perception of high and low probabilities towards certainty, and shifting the value of gains and losses such that large gains are progressively less significant, and losses more so. 
The wider implication is that social information flow may be distorted at the source, or through interpretation, and that agent-based models are uniquely suited to capture this.  We have focused here on descriptive theories of decision making, precisely because the aim is to model human behaviours as as they are manifested, rather than how they should be. Other approaches to decision models are available, for example heuristic methods \citep{Gigerenzer1996}, which emphasise the bounded rationality of human decision making, and Bayesian approaches, which address learning as a component of the decision process \citep{Robbins1964}. In an agent-based modelling context, \citet{Gray2016} have contrasted Bayesian, heuristic, and Cumulative Prospect Theory approaches in a simulation of alcohol misuse disclosure among pregnant women.

@@

Time is explicitly represented in almost all agent-based models, either in discrete or (pseudo) continuous form, in order to allow the dynamic unfolding of the modelled processes by `stepping' from one time-step to the next (or between time-ordered events in continuous-time discrete-event simulation). However, time often  does not enter into the decision making process of the simulated individuals. This results in the exclusion of a number of factors that may potentially be significant in eventual model outcomes. Firstly, individuals do not always make decisions instantaneously; as \cite{Willekens2017} describe elsewhere in this special issue, the Theory of Planned Behaviour \citep{Ajzen1991} can be used to describe the gradual process of intention forming, influence-absorption, information gathering and planning that terminate in a migration event. This drawn-out decision making process allows the possibility that agents may change their minds. Secondly, past experiences affect decisions in the present. This brings up questions about how the passage of time affects memory and attitudes. Finally, agents differ in their preferences between rewards realised immediately and equivalent rewards realised at some point in the future.

@@

Unfortunately, this is not the limit of potential complications around time. A large number of other contextual factors make a difference to the extent of discounting. For instance, it has been found that smaller rewards are more heavily discounted than larger rewards; that people treat a reductions in time to rewards differently to equivalent delays; and future losses treated as more significant than equivalent gains \citep{Read2000}.

@@

Uncertainty is a pervasive feature of any social, and thus also demographic reality. This uncertainty stems both from a lack of knowledge about processes and phenomena (\textit{epistemic uncertainty}), as well as the inherent randomness of the world (\textit{aleatory uncertainty} -- see e.g. \citeauthor{OHagan2004} \citeyear{OHagan2004}). This means that social actors make decisions under conditions of imperfect knowledge, but also that the reality is random, unpredictable and can yield many surprises. In agent-based modelling, these features need to be reflected in the simluated representations of the world, both with respect to agents themselves, as well as features of their environment.

As discussed in Section \ref{sec:context_matters}, human decisions are made under the conditions of bounded rationality and loss aversion. In the context of uncertainty, risk aversion is equally important. The underpinning theory, dating back at least to \citet{Pratt1964} and \citet{Arrow1965}, looks at the departures from the expected utility decision framework, and stipulates that risk-averse decision makers are willing to choose options with lower expected utility as long as they involve less, or no risk. Both aleatory and epistemic uncertainty are important here: through learning, agents can try reducing the epistemic uncertainty, but especially in the context of time-dependent agent based models, some of the uncertainty is irreducible, especially with respect to the decisions about future. This constitutes one of the main arguments for the importance of time, as argued in Section \ref{time}. Besides, varying attitudes to risk are an important source of heterogeneity of the agents, discussed in more detail in Section \ref{heterogeneity-in-decision-strategies}.

In the context of agent-based modelling of migration -- itself one of the most uncertain demographic processes -- there are some aspects where uncertainty just asks for being included in the models. For example, the uncertain nature of future benefits from migration, an important driver of population flows, has already been reflected in the neoclassical theories of migration, whereby earnings are weighted by the probability of employment, and future income streams are discounted to reflect the inter-temporal nature of decision making \citep{Massey1993}. In the New Economic Theory of migration, uncertainty comes to the fore of the decision-making process at the family or household level, and migration becomes one of the important tools of managing labour market risks \citep{Stark1985}. Uncertainty is at its most acute when we consider migration caused by extreme events, such as armed conflict or environmental catastrophes -- under such circumstances, decisions need to be made rapidly, sometimes with a very limited insight into available options and the potential trade-offs between costs, benefits and risks.

At the level of model, there are different layers of uncertainty: in the parameters, model specification, computer code, observations, not to mention the inherent residual variability \citep{Kennedy2001}. There are different ways to treat these manifestations of uncertainty in agent-based models. The use of statistical emulators or meta-models \citep{Kleijnen2000} -- statistical models approximating the key uncertain relationships between the parameters of the underlying complex computational models -- has been suggested by \citet{Kennedy2001} and \citet{Oakley2004}, and more recently, specifically in the context of agent-based modelling, by \citet{Heard2015} and \citet{Hilton2016}. Alternatives include the use of approximate Bayesian computation, as advocated by \citep{Grazzini2017}, or hybrid approaches, whereby emulators are enhanced by direct samples from the output of the underlying agent-based models \citep{Kaminski2015}.

All these different approaches aim to describe uncertainty in complex models in a coherent way, and to allow meaningful (and ideally computationally efficient) inference from observed data. The methodological developments in this area are likely to continue: \citet{Heard2015} noted a need to further enhance the statistical methods specifically designed for dealing with agent-based models. However, even now the available tools enable a statistically rigorous -- if at times approximate, as in the case of emulators -- analysis of uncertainty in selected aspects of the models at hand. Coupled with the explicit acknowledgement of uncertainty in human decision making, an honest account of the aleatory and epistemic limitations of agent-based models seems indispensable for their further methodological advancement and uptake, in demography and beyond.

@@

Second, the existing methods of the statistical design of experiments (e.g. \citeauthor{Chaloner} \citeyear{Chaloner}, \citeauthor{Kleijnen2000} \citeyear{Kleijnen2000}) would allow for adopting a systematic approach to model design and data collection. In this way, following the suggestions of \citet{Courgeau}, the simulations would be built iteratively, by applying the experimental design principles to identifying the elements of the model which require additional empirical information, and then by collecting this information.  In this way, the important aspects of the decision process discussed above – context, time and heterogeneity – can be explicitly built into the model, through its specification or parameters. Then, they can be learnt about through repeated collection of information – including through experiments. The modularity of the model would help this process, as it would allow for fine-tuning the different aspects of the model independently, before they can be assembled together. An outline of the process is illustrated in Figure \ref{fig:approach}.

@@

How might this work in practice? We recommend to begin the model building process by identifying the key features and the context of the phenomena being modelled. In the example of migration, the key aspects of the model would include not only the different types of heterogeneous agents -- migrants, non-migrants, institutions, intermediaries, policy makers, and so on -- but also the available contextual information: the known and relevant features of the geographic, social, and economic environments in which the agents operate. The selection of these features would be informed by the theoretical literature on migration (see \citeauthor{Massey1993} \citeyear{Massey1993} for examples), and to the greatest degree possible made realistic by including the empirical information -- quantitative as well as qualitative -- on what is known about migrant decision processes. The staged nature of decisions, as stipulated by the Theory of Planned Behaviour, could also be incorporated at this stage.

The second step of the model-buidling process would involve constructing a prototype, where the features that cannot be benchmarked to real-world data would be parameterised. Specifically in the context of decision making, such parameters could describe preferences related to risk or utility, aspects of decisions related to uncertainty or imperfect knowledge, time discounting. The choice of a model of decision making could be also parameterised: there may be many candidate frameworks (such as various heuristics, the Cumulative Prospect Theory or models of Bayesian reasoning -- see \citeauthor{Gray2016} \citeyear{Gray2016}), each with their idiosyncratic features, that can apply to different situations or types of agents. Crucially, for the purpose of constructing the agent-based models, all the unknown parameters would need to be described in probabilistic terms, and embedded within a framework of statistical design of experiments \citet{Chaloner}. In this way, the prototype agent-based model could be executed for different combinations of parameters, yielding different outputs.

The third step of the process would consist of calibrating the outputs of the models against the selected aspects of social reality, ideally in a probabilistic manner \citep{Hilton2016}. Once the model is calibrated, its statistical properties can be analysed. In particular, sensitivity of the outputs to the different parameters can be assessed \citep{Oakley2004} -- and the parameters that do not influence the model behaviour can be removed for the sake of parsimony. Conversely, we may want to find the empirical basis for those parameters, which appear to be highly influential for the chosen outputs, for example by collecting new data on the specific aspects of the agent-based model.

In particular, by calibrating the model, we can learn the choice of an appropriate representation of the decision processes of different agents. If the results of calibration do not point out to choosing a specific set of decision rules and models for various agents, this aspect of modelling can be enhanced by conducting cognitive experiments under controlled conditions. The sensitivity analysis would help identify, which aspects of the decision making process are the most important from the point of view of generating the observed outcomes -- this could be the treatment of time, uncertainty, or other parameters. In the fourth step of the process, the decision problems faced by the agents can be reproduced under laboratory conditions in a series of tasks to be solved by human volunteers participating in such experiments. The additional insights from this exercise would enable fine-tuning and re-parameterising of the agent-based model in the light of experimental findings, as proposed in the Manifesto of computational social science \citep{Conte}. The second, third and fourth steps above can then be iterated until no more improvements can be reasonably made -- what will be left is the residual uncertainty that is an irreducible feature of all models, and especially such complex ones.

Several examples of agent-based models presented in this Special Issue already contain some of the building blocks of the proposed approach. The multi-stage nature of the decision making processes in the context of migration has been explored by \citet{Willekens2017} and \citet{Kley2017}, both drawing from the Theory of Planned Behaviour \citep{Ajzen1991} and grounded in available empirical information. In addition, \citet{Warnke2017} have provided a stochastic description of the underlying decision making processes, and have proposed a corresponding programming language designed to facilitate the implementation and execution of agent-based models. The natural next step would be to use these building blocks, and others -- such as the experimental design, cognitive psychological experiments on decision making, or statistical analysis of uncertainty -- for constructing more robust and empirically grounded models of human decisions, which could then be embedded within agent-based models of human populations.

As argued throughout the Special Issue, all these features, and more, make agent-based modelling a very attractive analytical proposition for demographers. The potential gains in understanding of the underlying population processes cannot be overstated, and further links connecting agent-based and statistical demography are yet to be explored (see for example \citet{Bijak2016}). That at present there is no methodological consensus on modelling agency and human decisions is not an obstacle, but rather a challenge to guide the further work in this area. Given that over the last 15 years agent-based models have slowly begun to enter the demographic mainstream \citep{Billari2003,VanBavel2016}, the gaps in knowledge in the existing approaches have become clearly visible. In our view one of the important gaps is related to the agency and decision making processes of the simulated agents, and the related aspects, some of which – context, time, and a variety of forms of decision making – are discussed above. Now is the time to initiate the discussion about choosing the choice.
